OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-04-14 paper/paper.Rmd
Incorporate Pat's minor edits to results

- **start paragraph with question, end with why we should care. transitions to move the story along.**
- **AND, BUT, THEREFORE, structure**

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Initially, query sequences are placed in singleton OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the Matthews correlation coefficient (MCC). 
A sequence will remain unassigned if the MCC value is maximized when the
sequence is a singleton rather than assigned to a reference OTU.
This process is repeated until the MCC stabilizes, changing by no more than
0.0001 by default, or until a maximum number of iterations is reached (default:
100).
In closed-reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. 
The final quality score is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce stable OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs; the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3%.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
dropped to `r closed_fit_rdp_mcc`.
For comparison to the state of the art, clustering datasets _de novo_ with
OptiClust produced an average MCC score of `r opticlust_mcc`.
This gap in OTU quality mostly disappears when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
`r open_fit_silva_mcc` with Silva, and `r open_fit_rdp_mcc` with RDP.
Thus, open reference OptiFit produces OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit follows closely behind as long as a
suitable reference database is chosen.

Since closed reference clustering throws out query sequences that cannot be
assigned to reference OTUs, an additional measure of clustering performance to
consider is the fraction of query sequences that were able to be assigned.
On average, more sequences were assigned with Greengenes as the reference 
(`r frac_fit_gg`%) than with Silva (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
This mirrors the result reported above that Greengenes produced better OTUs in
terms of MCC score than either Silva or RDP.
Note that _De novo_ and open reference clustering methods always assign 100% of
sequences to OTUs.
The database chosen affects the final OTU assignments considerably in terms of 
both MCC score and fraction of query sequences fit.

Despite the drawbacks, closed reference methods have been used when fast
execution speed is required such as when using very large datasets.
To compare performance in terms of speed, we repeated each OptiFit and OptiClust
run 100 times and measured the execution time.
Closed reference OptiFit outperformed both OptiClust and open reference OptiFit,
with respective average run times of `r closed_fit_db_sec`, `r opticlust_sec`,
and `r open_fit_db_sec` seconds.
Thus, the OptiFit algorithm continues the precedent that closed reference
clustering sacrifices OTU quality for execution speed.

To compare to the reference clustering method used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
sequence's similarity to the single reference OTU.
In contrast, OptiFit accepts reference OTUs which each contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assign sequences to OTUs.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.
In closed reference mode, VSEARCH was able to map `r frac_vsearch_vs_fit`% more
query sequences than OptiFit to the Greengenes reference database.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc` (respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average respectively.
**TODO: conclude:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...

@@

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTUs quality may be poor when the reference database is too unrelated to the
samples of interest, such as when samples contain low abundant or rapidly
evolving microbes.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not stable when new sequences are
added.
Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
across studies or when making predictions with machine learning models.
To determine how well OptiFit performs for fitting new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.

>>>>

- **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
- From Pat: "briefly looking through the Discussion and Intro, one point that we may have forgotten is that a benefit of our approach is that it is much easier to customize to a specific region. The greengenes reference OTUs are based on full length sequences. This causes problems when considering shorter (e.g. V4) sequences since reference OTUs may be more similar  than the full length and even identical to each other in the subregion. Because we can easily spin up a subregion specific set of reference OTUs from a public database or the reference fraction this isn't a problem. This is described in one of my earlier papers looking at open/closed reference clustering and was part of the reason that the order of the database was important."

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
**TODO: brief description of the opticlust algorithm.**
OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Initially, query sequences are placed in singleton OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the Matthews correlation coefficient (MCC). 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.
**TODO: describe data sets here**

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce consistent OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3%.
The _de novo_ MCC scores for the three databases were **TODO**.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
dropped to `r closed_fit_rdp_mcc` when fitting to RDP.
For comparison, clustering datasets _de novo_ with
OptiClust produced an average MCC score of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

Since closed reference clustering does not cluster query sequences that could
not be assigned to reference OTUs, an additional measure of clustering
performance to consider is the fraction of query sequences that were able to be
assigned.
On average, more sequences were assigned with Greengenes as the reference 
(`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
This mirrored the result reported above that Greengenes produced better OTUs in
terms of MCC score than either SILVA or RDP.
Note that _de novo_ and open reference clustering methods always assign 100% of
sequences to OTUs.
The database chosen affects the final OTU assignments considerably in terms of
both MCC score and fraction of query sequences that could be fit to the
reference OTUs.

Despite the drawbacks, closed reference methods have been used when fast
execution speed is required such as when using very large datasets.
To compare performance in terms of speed, we repeated each OptiFit and OptiClust
run 100 times and measured the execution time.
Closed reference OptiFit outperformed both OptiClust and open reference OptiFit,
with average run times of `r closed_fit_db_sec`, `r opticlust_sec`,
and `r open_fit_db_sec` seconds, respectively.
**TODO: don't average by all datasets & datasets?.**
Thus, the OptiFit algorithm continues the precedent that closed reference
clustering sacrifices OTU quality for execution speed.

To compare to the reference clustering method used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
sequence's similarity to the single reference OTU.
In contrast, OptiFit accepts reference OTUs, which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assign sequences to OTUs.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.
In closed reference mode, VSEARCH was able to map `r frac_vsearch_vs_fit`% more
query sequences than OptiFit to the Greengenes reference database.
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of sequences in an OTU. <!-- VSEARCH "diameter" vs OptiFit "radius"
-->
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc` (respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average respectively.
**TODO: conclude:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...

@@

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTU quality may be poor when the reference database is too unrelated to the
samples of interest, such as when samples contain low abundant or novel
populations.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not consistent when new sequences are
added.
Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
across studies or when making predictions with machine learning models.
To determine how well OptiFit performs for fitting new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.
